---
title: "Assignment 5"
author: "Roland Abi"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 5)
```

```{r include=FALSE}
library(tidyverse)
library(lme4)
library(foreign)
library(arm)
library(data.table)
```

### Data preparation

```{r}
dta1<- read_csv("speed-1.dating.csv")

dta2_b <- dta1 |>
  dplyr::select(id, iid, partner, pid, dec, attr:shar) |>
  drop_na()
```

```{r}

n <- length(dta2_b$dec)
uniq <- unique(dta2_b$id)
ids <- rep(NA, n)

for (i in 1:length(uniq)) {
  ids[dta2_b$id == uniq[i]] <- i
}
dta2_b$ids <- ids
```

### Q14.5a

```{r}

complete_pooling <- glm(dec ~ attr + sinc + intel + fun + amb + 
                          shar, data = dta2_b, family = binomial(link = "logit"))

summary(complete_pooling)
```

$Attractiveness:$ the coefficient of `Attractiveness` is
`r coef(complete_pooling)[[2]]`, which when divided by 4 gives
`r (coef(complete_pooling)[[2]]/4)*100`, implies a 1% increase in attractiveness
score suggest that the rater was 15.9%% more likely to see their date
again.

$Sincere:$ the coefficient of `sincerity` is `r coef(complete_pooling)[[3]]`, which
when divided by 4 gives `r (coef(complete_pooling)[[3]]/4)*100`, implying a 1%
increase in sincerity score suggest that the rate was 1.74% more likely
to see their date again

$Inteligence:$ the coefficient of `Intelligence` is `r coef(complete_pooling)[[4]]`
which when divided by 4 gives `r (coef(complete_pooling)[[4]]/4)*100`, implies a 1%
increase in intelligence score suggest that the rater was 0.38% more
likely to see their date again.

$Fun:$ the coefficient of `fun` is `r coef(complete_pooling)[[5]]` which when divided
by 4 gives `r (coef(complete_pooling)[[5]]/4)*100`, which implies the funnier the
date, the more likely the rater will go on a repeat date.

$Ambitious:$ the coefficient of `Ambitious` is `r coef(complete_pooling)[[6]]` which
when divided by 4 gives `r (coef(complete_pooling)[[6]]/4)*100`, implies the less
ambitious the date, the less likely there is going to be a repeat date

$Shared interest:$ the coefficient of `shared interest` is
`r coef(complete_pooling)[[7]]` which when divided by 4 gives
`r (coef(complete_pooling)[[7]]/4)*100`, implies the chances of a repeat date
increases by 6.91% if the rater and their date share similar interests.

The The standard errors of the six variables were fairly within a
consistent range while the variables attractiveness, fun, ambitious and
shared interests, significantly contributed to determining if the rater
will be interested in seeing their date again.

### Q14.5b

```{r}
m2 <- glmer(dec ~ attr + sinc + intel + fun + amb + shar + (1|id), 
            data = dta2_b, family = binomial(link = "logit"),
           glmerControl(optimizer = c("bobyqa","Nelder_Mead"), 
                        optCtrl=list(maxfun=1e+05)))

summary(m2)
```

Model 2 (m2) which expands model (m1) to allow varying intercepts for
persons making the evaluation shows slightly similar coefficient
estimates. Having a group standard deviation 0.404 which is less than
the standard deviation for for the classical logistic regress model 1
(m1) `r sigma(complete_pooling)`

### Q14.5c

```{r}
m3 <- glmer(dec ~ attr + sinc + intel + fun + amb + shar + (1|id) + (1|pid), 
            data = dta2_b, family = binomial(link = "logit"),
           glmerControl(optimizer = c("bobyqa","Nelder_Mead"), 
                        optCtrl=list(maxfun=1e+05)))

summary(m3)
```

Model 3 (m3) which expands model 2 (m2) to allow varying intercepts for
persons making the evaluation shows slightly similar coefficient
estimates to model 2. Adding a variable for partner variability didn't
lead to a significant change in variance between between model 2 which
shows variability for persons making the evaluation

### Q14.6a

```{r}
no_pooling <- glm(dec ~ attr + sinc + intel + fun + amb + shar + as.factor(id)-1, data = dta2_b, 
          family = binomial(link = "logit"))

summary(no_pooling)
```

### Q14.6b

```{r}
m5 <- glmer(dec ~ scale(attr) + scale(sinc) + scale(intel) + scale(fun) + scale(amb)
           + scale(shar) + (scale(attr) + scale(sinc) + scale(intel) + scale(fun)
                            + scale(amb) + scale(shar) + 1|id), data = dta2_b, 
           family = binomial(link = "logit"))

summary(m5)
```

###Q14.6c

```{r}
anova(m5, complete_pooling, no_pooling)
```

The model evaluation shows that the multilevel model with varying
intercepts and coefficients performs better than the no pooling and
complete pooling models as showing in having lower AIC and deviance.

### Q15

```{r}
nes5200 <- read.dta("nes5200_processed_voters_realideo-1.dta")
```

### Q15.

```{r}
nes5200_dt <- data.table(nes5200)
  yr <- 2000
nes5200_dt_s<-nes5200_dt[ year==yr,]
nes5200_dt_s$income <- droplevels(nes5200_dt_s$income)
nes5200_dt_s$partyid7 <- droplevels(nes5200_dt_s$partyid7)

nes5200_dt_s$gender <- factor(nes5200_dt_s$gender, labels=c("male", "female"))
nes5200_dt_s$race <- factor(nes5200_dt_s$race, labels=c("white", "black", "asian", 
                                    "native american", "hispanic"))
nes5200_dt_s$south <- factor(nes5200_dt_s$south)
nes5200_dt_s$ideo <- factor(nes5200_dt_s$ideo, labels=c("liberal", "moderate", "conservative"))

nes_data_comp<-nes5200_dt_s[complete.cases(nes5200_dt_s[,list(partyid7,income,ideo,female,white)])]
nes_data_comp$ideology <- scale(nes_data_comp$ideo_feel,center=TRUE)
```

```{r}
newnes5200 <- nes5200_dt_s |>
  mutate(
    new_partyid7 = as.numeric(str_extract(partyid7, "[[:digit:]]"))
  ) |> relocate(new_partyid7)

nes_5200 <- newnes5200[complete.cases(newnes5200, list(new_partyid7, gender, race, ideo))]

new
```

### Q15.1

$yij$ =
$\begin{cases} 0 & if & z_{i} & = & Gore \\ 1 & if & z_{i}&=&noopinion \\ 2&if&z_{i}&=&Bush \end{cases}$

$z_i \sim logistic(X_i\beta, \sigma^2)$

$z_i = logit^-1(\alpha_{i[j]} + partyideology_{kj[i]} + ethnicity_{lj[i]} + Age_{mj[i]} + Education_{nj[i]})$

$where$ $\alpha_j \sim N(0, \delta^2_\alpha)$ $j = 1,\dots ,nstates$

### Q15.2a

$yij$ =
$\begin{cases} 1 & if & z_{i} & = & strong-democrat \\ 2 & if & z_{i}&=&weak-democrat \\ 3&if&z_{i}&=&independent-democrat \\ 4&if&z_{i}&=&independent-independent \\ 5&if&z_{i}&=&independent-republican \\ 6&if&z_{i}&=&weak-republican \\ 7&if&z_{i}&=&strong-republican \end{cases}$

$z_i \sim logistic(X_i\beta, \sigma^2)$

$z_i = logit^-1(\alpha_{i[j]} + partyideology_{kj[i]} + ethnicity_{l} + Age_{m} + gender_{n})$

$k = 1, \dots,3 l = 1,\dots,4; n = 1, \dots,2; j = 1,\dots ,nstates; i = 1,\dots,n$

$where$ $\alpha_j \sim N(0, \delta^2_\alpha)$

### Q15.2b

```{r}
m6 <- lmer(scale(new_partyid7) ~ ideo + scale(black) + scale(female) + scale(age) +  
             (1 + ideo|state), data = newnes5200)
```
